Appendices

### 2022-10-02 appendices/are_the_series_purely_random.Rmd
correct model specification for arima(0,1,0). Update summary of results accordingly

---
title: "Time Series Models"
author: "Jon Minton"
date: '2022-07-21'
output:
  pdf_document: default
  html_document: default
---

@@

# Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?

Section X of the main manuscript noted, from figure 1a and 1b, that there is substantial annual variation in annual changes in life expectancy. After taking the first difference in these series (i.e. subtracting for each value in the series its previous value), the series of remaining difference may be purely random. Alternatively, even after such differencing the series may have some degree of autocorrelation (i.e. the differenced value for one year may be informative as to the differenced value for the next one or more observations in the series), or there may be a moving average component to the series.

@@

ts_model_comparisons <- 
  series_of_interest %>% 
    group_by(country, sex, x) %>% 
    nest() %>% 
    mutate(
      arima_010 = map(data, fit_arima_model, order = c(1, 1, 0)),
      arima_110 = map(data, fit_arima_model, order = c(1, 1, 0))
    ) %>% 
  mutate(
    aicc_arima_010 = map_dbl(arima_010, ~summary(.) %>% pluck("aicc")),
    aicc_arima_110 = map_dbl(arima_110, ~summary(.) %>% pluck("aicc"))
  ) %>% 
  mutate(
    which_preferred = if_else(aicc_arima_010 < aicc_arima_110, "Random Walk", "Autocorrelated")
  )

@@

The Autocorrelated specification (ARIMA(1,1,0)) is preferred to RWD (ARIMA(0,1,0)) for all populations.

@@

## Alternative approach to establishing if series are autocorrelated

An alternative way to determing whether the series are autocorrelated is to compare two linear regression equations: one in which change in life expectancy is regressed against an intercept; another in which the change in life expectancy is additionally regressed against the change in the previous year

alt_approach_autocorrelated <- 
  hmd_ex_selected_countries_with_synth %>% 
      filter(year >= 1979) %>% 
      filter(!(code %in% c("DEUTE", "DEUTW", "DEUTNP"))) %>%  # Using only synthetic germany for longer time period
      group_by(code, x, sex) %>% 
      arrange(year) %>% 
      mutate(delta_ex = ex - lag(ex)) %>% 
      nest() %>% 
      mutate(
        rwd_model = map(data, ~lm(delta_ex ~ 1, data = . )),
        ar_model  = map(data, ~lm(delta_ex ~ lag(delta_ex), data = .))
      ) %>% 
      mutate(
        bic_rwd = map_dbl(rwd_model, BIC),
        bic_ar  = map_dbl(ar_model, BIC)
      ) %>% 
      mutate(
        diff_bic = bic_ar - bic_rwd
      ) %>% 
    mutate(
      which_preferred = ifelse(diff_bic < 0, "Autocorrelated", "RWD")
    ) %>% 
      left_join(country_code_lookup) %>% 
      mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
    filter(!is.na(country))

alt_approach_autocorrelated %>% 
  select(country, x, sex, bic_rwd, bic_ar, which_preferred)
```
This approach also identifies most, but not all, of the populations to exhibit some degree of autocorrelation. (n.b. the more stringent BIC was used instead of AICc in this example, which may explain discrepancies between approaches)

For each of the above, the intercept (average improvement per year) and autocorrelation term can be extracted as follows:

```{r}
alt_approach_autocorrelated  %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  select(country, x, sex, ar_model) %>% 
  mutate(tidied_model = map(ar_model, broom::tidy)) %>% 
  select(-ar_model) %>% 
  filter(x == 0) %>% 
  unnest(tidied_model) %>% 
  ggplot(aes(estimate, reorder(country, estimate))) + 
  geom_point() + 
  facet_grid(term~sex) + 
  geom_vline(xintercept = 0) + 
  geom_errorbarh(aes(xmin = estimate - 1.96  * std.error, xmax = estimate + 1.96 * std.error))  + 
  labs(
    x = "Coefficient value", 
    y = "Country",
    title = "Coefficients from autoregressive models\nlife expectancy at birth, selected countries",
    subtitle = "Horizontal bars indicate 95% confidence intervals",
    caption = "Source: HMD lifetables"

  )

```
Here all autoregression coefficients are negative, most of which are statistically significant at p < 0.05.

The following shows the equivalent trends for x = 65

```{r}
alt_approach_autocorrelated  %>% 
  left_join(country_code_lookup) %>% 
  mutate(country = factor(country, levels = c("England & Wales", "Scotland", "Synthetic Germany", "Spain", "France", "Italy", "Netherlands"))) %>% 
  filter(!is.na(country)) %>% 
  select(country, x, sex, ar_model) %>% 
  mutate(tidied_model = map(ar_model, broom::tidy)) %>% 
  select(-ar_model) %>% 
  filter(x == 65) %>% 
  unnest(tidied_model) %>% 
  ggplot(aes(estimate, reorder(country, estimate))) + 
  geom_point() + 
  facet_grid(term~sex) + 
  geom_vline(xintercept = 0) + 
  geom_errorbarh(aes(xmin = estimate - 1.96  * std.error, xmax = estimate + 1.96 * std.error))  + 
  labs(
    x = "Coefficient value", 
    y = "Country",
    title = "Coefficients from autoregressive models\nlife expectancy at age 65, selected countries",
    subtitle = "Horizontal bars indicate 95% confidence intervals",
    caption = "Source: HMD lifetables"

  )
```
Here the negative correlations appear even stronger, and of larger magnitude in Italy than elsewhere. It also appears that the two coefficients are negatively associated, with Netherlands having the lowest average annual improvement, but also the smallest (and non-significant) magnitude of negative autocorrelation; Italy appears the converse.

@@

The length of the vectors ar_roots and ma_roots indicate, respectively, how many ar or ma terms were identified in the best fitting model for the population indicated by sex, x (starting age) and country. For example, for females in England & Wales, from age 0, an ARIMA(1,1,0) model is preferred, whereas for females in Scotland, from age 0, an ARIMA(0, 1, 1) model is preferred.

>>>>

---
title: "Appendices"
author: "Jon Minton"
date: '2022-07-21'
output: html_document
---

@@

# Appendix 1: Are annual ex time series random walks with drift (RWD) or do they have autocorrelation?

Our paper noted that there is substantial annual variation in annual changes in life expectancy. After taking the first difference in these series (i.e. subtracting for each value in the series its previous value), the series of remaining difference may be purely random. Alternatively, even after such differencing the series may have some degree of autocorrelation (i.e. the differenced value for one year may be informative as to the differenced value for the next one or more observations in the series), or there may be a moving average component to the series.

@@

ts_model_comparisons <- 
  series_of_interest %>% 
    group_by(country, sex, x) %>% 
    nest() %>% 
    mutate(
      arima_010 = map(data, fit_arima_model, order = c(0, 1, 0)), # corrected from c(1, 1, 0)
      arima_110 = map(data, fit_arima_model, order = c(1, 1, 0))
    ) %>% 
  mutate(
    aicc_arima_010 = map_dbl(arima_010, ~summary(.) %>% pluck("aicc")),
    aicc_arima_110 = map_dbl(arima_110, ~summary(.) %>% pluck("aicc"))
  ) %>% 
  mutate(
    which_preferred = if_else(aicc_arima_010 < aicc_arima_110, "Random Walk", "Autocorrelated")
  )

@@

For females, the Random Walk with Drift (RWD, ARIMA(0, 1, 0)) specification is preferred for all populations considered except Scottish and German females from age 65.

For males, the Autocorrelated specification (ARIMA(1, 1, 0)) is preferred for all populations for changes in life expectancy from birth (x = 0), whereas from age 65 the RWD specification is preferred for Spain, France, Italy,  and Scotland.

@@

@@

The length of the vectors ar_roots and ma_roots indicate, respectively, how many ar or ma terms were identified in the best fitting model for the population indicated by sex, x (starting age) and country. For example, for females in England & Wales, from age 0, an ARIMA(1,1,0) model is preferred, whereas for females in Scotland, from age 0, an ARIMA(0,1,1) model is preferred.
